\chapter{Test Setup}
\label{chap:Test Setup}
We have created some test cases.

\section{Memory Usage in DataMart}
\label{sec:Memory Usage in DataMart}

\subsection{Windows Memory Manager}
\label{sub:Windows Memory Manager}
There is much overhead in the applications, especially in .NET.

We do have accurate data on the memory used by the delphi layer of the code. However, from an end user perspective, the memory is measured in the windows memory manager. We have, therefore, used this tool to measure the memory footprint of our application.


\section{Real customer test cases}
\label{sec:Real customer test cases}
We also run some tests on some real test cases proposed by \genus.

Both test cases will be run and presented to the respective customers.

Due to confidential information, the names and the specific product are not available in this report, and the respective companies are held anonymous.

\subsection{Real Customer Test Case I: A Stock Broker}
\label{sub:Real Customer Test Case I: A Stock Broker}
The first test case tests memory usage and load time for an analysis on stock ownership.
\begin{itemize}
    \item Distributor and Customer Reporting
    \item Operations, Settlements and Fund Management Support
    \item Payments and Payment matching
    \item Commissions Management
    \item Customer Relationship Management (CRM)
    \item Portfolio and Company Management
    \item Order processing
\end{itemize}

\subsection{Real Customer Data Test Case II: A Retail Chain}
\label{sub:Real Customer Data Test Case II: A Retail Chain}
\begin{itemize}
    \item Manage operations for 2200 grocery stores
    \item Corporate Performance Management (CPM)
    \item Apps for Operational Control \& Monitoring
    \item Employee- and customer surveys
    \item Food Safety and Risk Management
\end{itemize}
Like above, this is tested for memory usage and load time.



