{\centering

    {\Huge Preface}
    \vspace{1cm}
    % The preface includes the facts - what type of project, where it is conducted, who supervised and any acknowledgements.
    % http://www.writersandeditors.com/preface__foreword__or_introduction__57375.htm

    After a long fall, we finally concluded a paper on database techonolgies used for business intelligence. However, we came across with an idea that was rather controversial. Why not implement column store within the core of \genusSoftware?

    As a collaboration project between \genus~and \ntnu~to improve performance, we implement column storage within the core of \genusSoftware. The research in mainly conducted by \textit{Håkon Åmdal} with the assistance and supervision of \textit{Einar Bleikvin} and \supervisor.
    \vspace{1cm}

    {\large Acknowledgements}

    First of all, I would like to thank \genus~for proposing an interesting and challenging research project. It has been compelling working with in-memory, \bd~products, but most importantly; working on a real business case.

    It is my pleasure to express my gratitude to \textit{Einar Bleikvin}, my co-supervisor at \genus. He has during the project period always shown interest in my work, and he has been a great discussion partner. Einar is not easily convinced, which have motivated me to do a thorough job before presenting my findings to him. I look forward to working closer with Einar in the spring project.

    \textit{Johnny Troset Andersen} deserves recognition for his dedication to this research. Johnny has always provided be with the resources I need to execute the project, which includes access to key personnel within \genus~and relevant literature. Last, but not least, he has always provided me with a desk at \genus' office and made me feel like home.

    Other \genus~employees that deserve recognition include \textit{Thomas Hedén}, which demonstrated \qlikview~to me and helped me access customer data, \textit{Bernt Almlid} that held a lesson in \genusSoftware~multitenancy and architecture, and \textit{Martin Børke} for input on the history of \bi. These people have also provided valuable feedback on the project report.

    I would like to thank \textit{Svein Erik Bratsberg} for being my supervisor on this project. Svein Erik has kept me on track for the project period and has always held his office door open for me. He has provided me with key literature for this project, and provided valuable feedback on the project report. Svein Erik has been an irreplaceable asset for this project, especially in the final couple of weeks.

    Lastly, I am extremely grateful for \textit{Marte Løge}'s support.

}
